\texttt{methyvim}: Targeted and Model-free Differential Methylation Analysis in R

### 2018-04-16 paper/paper_BiocF1000.tex
F1000 (#34)
* use minfi data

* add paper refs

* started to add a bit to the introduction &  moved the text from the repository to the appropriate sections in the article. I will use the text from the repo as a rough guide to draft something more cohesive

* revisions/updates to F1000 manuscript (#28)

* minor vignette additions/corrections

* almost done with f1000, minor vignette corrections

* adding the Summary and final touches

* a few more minor things

* fixing merge conflict

* minor formatting

* suggestion: convert betas to mvals in the limma screenas the mvals have far more centra

\usepackage[numbers]{natbib}

@@

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}

@@

@@

\title{\texttt{methyvim}: Targeted and Model-free Analysis of Differential
Methylation in R}

\author[1]{Nima S. Hejazi}
\author[1]{Rachael V. Phillips}
\author[1]{Alan E. Hubbard}
\author[1]{Mark J. van der Laan}
\affil[1]{Division of Biostatistics, University of California, Berkeley}

@@

\section{Introduction}\label{introduction}

DNA methylation is a fundamental epigenetic process known to play an
important role in the regulation of gene expression. DNA CpG methylation
involves the addition of a methyl group (\(\text{CH}_3\)) to the fifth
carbon of the cytosine ring structure to form 5-methylcytosine. Numerous
biological and medical studies have implicated DNA CpG methylation as
playing a role in disease and development (Robertson 2005). Perhaps
unsurprisingly then, biotechnologies have been developed to study the
molecular mechanisms of this epigenetic process. Modern assays, like the
Illumina \emph{Infinium} Methylation assay, allow for quantitative
interrogation of DNA methylation of CpG sites scattered across the
genome at single-nucleotide resolution; moreover, much effort has been
invested, by the bioinformatics community, in the development of tools
for properly removing technological effects that may contaminate
biological signatures measured by such assays. Despite these advances in
both biological and bioninformatical techniques, most statistical
methods available for differential analysis of data produced by such
assays rely on over-simplified (often generalized linear) models. This
is because typical statistical estimation approaches are not achievable
for such high-dimensional data without restrictive modeling assumptions.
When these assumptions are violated, estimates of the population-level
effect of an exposure or treatment may have large bias. Data-adaptive
estimation procedures that utilize machine learning can control for
high-dimensional confounders when estimating treatment/exposure effects.
However, interpretable statistical inference (i.e.~confidence intervals,
hypothesis tests) resulting from these data-adaptive estimates is
challenging (Libbrecht and Noble 2015).

In this paper, we present an alternative such statistical analysis
approaches in the form of nonparametric estimation procedures that
provide simple and well interpretable statistical inference. Inspired by
concepts of counterfactual effects from the causal inference literature
and machine learning algorithms, we provide a computationally efficient
technique for obtaining targeted estimates of nonparametric
\emph{variable importance measures} (\textbf{VIM}) for a pre-screened
set of CpGs, controlling for the False Discovery Rate (FDR) as if all
sites were tested. Under counterfactual assumptions, these estimators
have asymptotic normal sampling distributions and thus provide reliable
inference. Counterfactuals are ``counter''-fact, defined in terms of an
unrealistic yet ideal experiment. In the context of DNA methylation and
for a binary treatment, we define the counterfactual outcomes as the
observed methylation value (whether Beta- or M-values) a CpG site would
have if all subjects were treated and the methylation value a CpG site
would have if all subjects were not treated. Of course, these
counterfactual outcomes are impossible to observe but they are very
helpful for us to asses the importance of a CpG site relative to a
subjects treatment/exposure and they can be statistically estimated (M.
J. van der Laan 2006). Our VIM definition uses the concepts of
counterfactuals and the framework for estimation is built on
\emph{targeted maximum likelihood estimation} or \emph{targeted minimum
loss-based estimation} (\textbf{TMLE}) (M. J. van der Laan and Rubin
2006). This methodology targets the individual importance of each CpG
site by taking advantage of the prediction power of machine learning
algorithms. The mathematical properties of this approach have been
explored in (Tuglus and van der Laan 2008) and we focus on our
contribution to it's application, \texttt{methyvim}, an R package that
make this technique for differential methylation analysis easily
accessible.

For a general discussion of the framework of targeted minimum loss-based
estimation and the role this approach plays in statistical causal
inference, the interested reader is invited to consult M. J. van der
Laan and Rose (2011) and M. J. van der Laan and Rose (2017). For a more
general introduction to (statistical) causal inference, Pearl (2009) and
Hernan and Robins (2018, forthcoming) may be of interest.

\section{Methods}\label{methods}

\subsection{Implementation}\label{implementation}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Pre-screening of genomic sites is used to isolate a subset of sites
  for which there is cursory evidence of differential methylation. For
  the sake of computational feasibility, targeted minimum loss-based
  estimates of VIMs are computed only for this subset of sites.
  Currently, the available screening approach adapts core routines from
  the \href{http://bioconductor.org/packages/limma}{\texttt{limma}} R
  package. Future releases will support functionality from other
  packages (e.g.,
  \href{https://CRAN.R-project.org/package=randomForest}{\texttt{randomForest}},
  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}}).
  Following the style of the function for performing screening via
  \texttt{limma}, users may write their own screening functions and are
  invited to contribute such functions to the core software package by
  opening pull requests at the GitHub repository.
\item
  Nonparametric estimates of VIMs, for the specified target parameter,
  are computed at each of the CpG sites passing the screening step. The
  VIMs are defined in such a way that the estimated effects is of an
  exposure/treatment on the methylation status of a target CpG site,
  controlling for the observed methylation status of the neighbors of
  that site. Currently, routines are adapted from the
  \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R
  package. Future releases will support doubly-robust estimates of these
  VIMs (via the
  \href{https://cran.r-project.org/package=drtmle}{\texttt{drtmle}}
  package) and add parameters for continuous treatments/exposures (via
  the
  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}}
  package).
\item
  Since pre-screening is performed prior to estimating VIMs, we make use
  of a multiple testing correction uniquely suited to such settings. Due
  to the multiple testing nature of the estimation problem, a variant of
  the Benjamini \& Hochberg procedure for controlling the False
  Discovery Rate (FDR) is applied (Benjamini and Hochberg 1995).
  Specifically, we apply the ``modified marginal Benjamini \& Hochberg
  step-up False Discovery Rate controlling procedure for multi-stage
  analyses'' (FDR-MSA), which is guaranteed to control the FDR as if all
  sites were tested (i.e., without screening) (Tuglus and van der Laan
  2009).
\end{enumerate}

\paragraph{Parameters of Interest}\label{parameters-of-interest}

@@

Estimating the VIM corresponding to the parameters above, for
discrete-valued treatments or exposures, requires two separate
regression steps: one for the treatment mechanism (propensity score) and
one for the outcome regression. Technical details on the nature of these
regressions are discussed in Hernan and Robins (2018, forthcoming), and
details for estimating these regressions in the framework of targeted
minimum loss-based estimation are discussed in M. J. van der Laan and
Rose (2011).

@@

\begin{itemize}
\item
  A nonparametric variable importance measure (NPVI) (Chambaz, Neuvial,
  and van der Laan 2012): The effect of continuous-valued exposure or
  treatment (the observed methylation at a target CpG site) on an
  outcome of interest is estimated, controlling for the observed
  methylation at all other CpG sites in the same neighborhood as the
  target (treatment) site, based on a parameter that compares values of
  the treatment against a reference value taken to be the null. In
  particular, the implementation provided is designed to assess the
  effect of differential methylation at the target CpG site on a
  (typically) phenotype-level outcome of interest (e.g., survival), in
  effect providing an nonparametric evaluation of the impact of
  methylation at the target site on said outcome.
\end{itemize}

\paragraph{\texorpdfstring{Class
\texttt{methytmle}}{Class methytmle}}\label{class-methytmle}

@@

We refer the reader to the package vignette, ``\texttt{methyvim}:
Targeted Learning for Differential Methylation Analysis,'' included in
any distribution of the software package, for further details.

\subsection{Operation}\label{operation}

@@

\section{Use Cases}\label{use-cases}

@@

\paragraph{Preliminaries: Setting up the
Data}\label{preliminaries-setting-up-the-data}

@@

\begin{verbatim}
## methyvim: Targeted Learning for Differential Methylation Analysis
\end{verbatim}

\begin{verbatim}
## Version: 0.99.8
\end{verbatim}

@@

\paragraph{Differential Methylation
Analysis}\label{differential-methylation-analysis}

@@

\begin{verbatim}
##            lowerCI_ATE      est_ATE   upperCI_ATE      Var_ATE
## cg14008030 -0.11956597 -0.031415962  0.0567340489 2.022705e-03
## cg20253340 -0.08850637 -0.058866142 -0.0292259165 2.286919e-04
## cg21870274 -0.09499057 -0.029118982  0.0367526091 1.129495e-03
## cg17308840 -0.04626018 -0.007152452  0.0319552773 3.981191e-04
## cg00645010 -0.02677328 -0.013465554 -0.0001578256 4.609945e-05
## cg27534567  0.06745648  0.115711854  0.1639672225 6.061486e-04
##                    pval n_neighbors_all n_neighbors_w max_corr_w
## cg14008030 4.848468e-01               0             0         NA
## cg20253340 9.917426e-05               0             0         NA
## cg21870274 3.862537e-01               2             1  0.9443580
## cg17308840 7.199943e-01               2             1  0.9443580
## cg00645010 4.734007e-02               2             2  0.5236810
## cg27534567 2.602936e-06               1             0  0.9362968
\end{verbatim}

Finally, we may compute FDR-corrected p-values, by applying a modified
procedure for controlling the False Discovery Rate for multi-stage
analyses (FDR-MSA) (Tuglus and van der Laan 2009). We do this by simply
applying the \texttt{fdr\_msa} function.

@@

\paragraph{Visualization of Results}\label{visualization-of-results}

@@

\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-pvals-raw-1.pdf}

@@

\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-pvals-fdr-1.pdf}

@@

\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-volcano-1.pdf}

@@

\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-heatmap-1.pdf}

Invoking \texttt{methyheat} in this manner produces a plot of the top
sites (\(25\), by default) based on the raw p-value, using the raw
methylation measures in the plot. This uses the exceptional
\texttt{superheat} R package ({\textbf{???}}).

\section{Summary}\label{summary}

@@

\section{Software availability}\label{software-availability}

@@

\section{Author contributions}\label{author-contributions}

NH designed and implemented the software package, applied the tool to
the use cases, and co-drafted the present article. RP helped in
designing the software and co-drafted the present article. AH and ML
serve as the advisors for the development of this software and general
statistical algorithm.

\section{Competing interests}\label{competing-interests}

@@

\section{Grant information}\label{grant-information}

NH was supported in part by the National Library Of Medicine of the
National Institutes of Health under Award Number T32LM012417, by
{[}SUPERFUND GRANT{]}, and by {[}NINA HOLLAND GRANT{]}. RP was supported
by {[}SUPERFUND GRANT{]}. The content of this work is solely the
responsibility of the authors and does not necessarily represent the
official views of the various funding sources and agencies.

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\hypertarget{ref-benjamini1995controlling}{}
Benjamini, Yoav, and Yosef Hochberg. 1995. ``Controlling the False
Discovery Rate: A Practical and Powerful Approach to Multiple Testing.''
\emph{Journal of the Royal Statistical Society. Series B
(Methodological)}. JSTOR, 289--300.

\hypertarget{ref-chambaz2012estimation}{}
Chambaz, Antoine, Pierre Neuvial, and Mark J van der Laan. 2012.
``Estimation of a Non-Parametric Variable Importance Measure of a
Continuous Exposure.'' \emph{Electronic Journal of Statistics} 6. NIH
Public Access: 1059.

\hypertarget{ref-hernan2018causal}{}
Hernan, Miguel A, and James M Robins. 2018, forthcoming. \emph{Causal
Inference}. Chapman \& Hall/Crc Texts in Statistical Science. Taylor \&
Francis.

\hypertarget{ref-libbrecht2015machine}{}
Libbrecht, Maxwell W, and William Stafford Noble. 2015. ``Machine
Learning in Genetics and Genomics.'' \emph{Nature Reviews. Genetics} 16
(6). NIH Public Access: 321.

\hypertarget{ref-pearl2009causality}{}
Pearl, Judea. 2009. \emph{Causality: Models, Reasoning, and Inference}.
Cambridge University Press.

\hypertarget{ref-robertson2005dna}{}
Robertson, Keith D. 2005. ``DNA Methylation and Human Disease.''
\emph{Nature Reviews. Genetics} 6 (8). Nature Publishing Group: 597.

\hypertarget{ref-tuglus2008targeted}{}
Tuglus, Catherine, and Mark J van der Laan. 2008. ``Targeted Methods for
Biomarker Discovery, the Search for a Standard.'' bepress.

\hypertarget{ref-tuglus2009modified}{}
Tuglus, Catherine, and Mark J. van der Laan. 2009. ``Modified FDR
Controlling Procedure for Multi-Stage Analyses.'' \emph{Statistical
Applications in Genetics and Molecular Biology} 8 (1). Walter de
Gruyter: 1--15.
doi:\href{https://doi.org/10.2202/1544-6115.1397}{10.2202/1544-6115.1397}.

\hypertarget{ref-vdl2006statistical}{}
van der Laan, Mark J. 2006. ``Statistical Inference for Variable
Importance.'' \emph{The International Journal of Biostatistics} 2 (1).

\hypertarget{ref-vdl2011targeted}{}
van der Laan, Mark J, and Sherri Rose. 2011. \emph{Targeted Learning:
Causal Inference for Observational and Experimental Data}. Springer
Science \& Business Media.

\hypertarget{ref-vdl2017targeted}{}
---------. 2017. \emph{Targeted Learning in Data Science: Causal
Inference for Complex Longitudinal Studies}. Springer Science \&
Business Media.

\hypertarget{ref-vdl2006targeted}{}
van der Laan, Mark J, and Daniel Rubin. 2006. ``Targeted Maximum
Likelihood Learning.'' \emph{The International Journal of Biostatistics}
2 (1).

>>>>

\usepackage[numbers]{natbib}
\bibliographystyle{unsrtnat}

@@

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

@@

\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}

@@

\title{\texttt{methyvim}: Targeted and Model-free Differential Methylation Analysis in R}

\author[1,2]{Nima S. Hejazi\thanks{\ttfamily nhejazi@berkeley.edu}}
\author[1]{Rachael V. Phillips}
\author[1]{Alan E. Hubbard}
\author[1,3]{Mark J. van der Laan}
\affil[1]{Group in Biostatistics, University of California, Berkeley}
\affil[2]{Center for Computational Biology, University of California, Berkeley}
\affil[3]{Department of Statistics, University of California, Berkeley}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

DNA methylation is a fundamental epigenetic process known to play an important
role in the regulation of gene expression. DNA CpG methylation involves the
addition of a methyl group (\(\text{CH}_3\)) to the fifth carbon of the cytosine
ring structure to form 5-methylcytosine. Numerous biological and medical studies
have implicated DNA CpG methylation as playing a role in disease and
development \citep{robertson2005dna}. Perhaps unsurprisingly then, biotechnologies
have been developed to study the molecular mechanisms of this epigenetic
process. Modern assays, like the Illumina \emph{Infinium} Methylation assay,
allow for quantitative interrogation of DNA methylation of CpG sites scattered
across the genome at single-nucleotide resolution; moreover, much effort
has been invested, by the bioinformatics community, in the development of tools
for properly removing technological effects that may contaminate biological
signatures measured by such assays. Despite these advances in both biological
and bioninformatical techniques, most statistical methods available for
differential analysis of data produced by such assays rely on over-simplified
(often generalized linear) models. This is because typical statistical
estimation approaches are not achievable for such high-dimensional data without
restrictive modeling assumptions. When these assumptions are violated, estimates
of the population-level effect of an exposure or treatment may have large bias.
Data-adaptive estimation procedures that utilize machine learning can control
for high-dimensional confounders when estimating treatment/exposure effects.
However, interpretable statistical inference (i.e.~confidence intervals,
hypothesis tests) resulting from these data-adaptive estimates is challenging
\citep{libbrecht2015machine}.

In this paper, we present an alternative such statistical analysis approaches in
the form of nonparametric estimation procedures that provide simple and well
interpretable statistical inference. Inspired by concepts of counterfactual
effects from the causal inference literature and machine learning algorithms, we
provide a computationally efficient technique for obtaining targeted estimates
of nonparametric \emph{variable importance measures} (\textbf{VIM}) for a pre-screened set
of CpGs, controlling for the False Discovery Rate (FDR) as if all sites were
tested. Under counterfactual assumptions, these estimators have asymptotic
normal sampling distributions and thus provide reliable inference.
Counterfactuals are ``counter''-fact, defined in terms of an unrealistic yet ideal
experiment. In the context of DNA methylation and for a binary treatment, we
define the counterfactual outcomes as the observed methylation value (whether
Beta- or M-values) a CpG site would have if all subjects were treated and the
methylation value a CpG site would have if all subjects were not treated. Of
course, these counterfactual outcomes are impossible to observe but they are
very helpful for us to asses the importance of a CpG site relative to a subjects
treatment/exposure and they can be statistically estimated
\citep{vdl2006statistical}. Our VIM definition uses the concepts of counterfactuals
and the framework for estimation is built on \emph{targeted maximum likelihood
estimation} or \emph{targeted minimum loss-based estimation} (\textbf{TMLE})
\citep{vdl2006targeted}. This methodology targets the individual importance of each
CpG site by taking advantage of the prediction power of machine learning
algorithms. The mathematical properties of this approach have been explored in
\citep{tuglus2008targeted} and we focus on our contribution to it's application,
\texttt{methyvim}, an R package that make this technique for differential methylation
analysis easily accessible.

For a general discussion of the framework of targeted minimum loss-based
estimation and the role this approach plays in statistical causal inference, the
interested reader is invited to consult \citet{vdl2011targeted} and \citet{vdl2017targeted}.
For a more general introduction to (statistical) causal inference,
\citet{pearl2009causality} and \citet{hernan2018causal} may be of interest.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{implementation}{%
\subsection{Implementation}\label{implementation}}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Pre-screening of genomic sites is used to isolate a subset of sites for
  which there is cursory evidence of differential methylation. For the sake of
  computational feasibility, targeted minimum loss-based estimates of VIMs are
  computed only for this subset of sites. Currently, the available screening
  approach adapts core routines from the
  \href{http://bioconductor.org/packages/limma}{\texttt{limma}} R
  package. Future releases will support functionality from other packages
  (e.g., \href{https://CRAN.R-project.org/package=randomForest}{\texttt{randomForest}},
  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}}). Following the
  style of the function for performing screening via \texttt{limma}, users may write
  their own screening functions and are invited to contribute such functions to
  the core software package by opening pull requests at the GitHub repository.
\item
  Nonparametric estimates of VIMs, for the specified target parameter, are
  computed at each of the CpG sites passing the screening step. The VIMs are
  defined in such a way that the estimated effects is of an exposure/treatment
  on the methylation status of a target CpG site, controlling for the observed
  methylation status of the neighbors of that site. Currently, routines are
  adapted from the \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R package.
  Future releases will support doubly-robust estimates of these VIMs (via the
  \href{https://cran.r-project.org/package=drtmle}{\texttt{drtmle}} package)
  and add parameters for continuous treatments/exposures (via the
  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}} package).
\item
  Since pre-screening is performed prior to estimating VIMs, we make use of a
  multiple testing correction uniquely suited to such settings. Due to the
  multiple testing nature of the estimation problem, a variant of the Benjamini
  \& Hochberg procedure for controlling the False Discovery Rate (FDR) is
  applied \citep{benjamini1995controlling}. Specifically, we apply the ``modified
  marginal Benjamini \& Hochberg step-up False Discovery Rate controlling
  procedure for multi-stage analyses'' (FDR-MSA), which is guaranteed to
  control the FDR as if all sites were tested (i.e., without screening)
  \citep{tuglus2009modified}.
\end{enumerate}

\hypertarget{parameters-of-interest}{%
\paragraph{Parameters of Interest}\label{parameters-of-interest}}

@@

Estimating the VIM corresponding to the parameters above, for discrete-valued
treatments or exposures, requires two separate regression steps: one for the
treatment mechanism (propensity score) and one for the outcome regression.
Technical details on the nature of these regressions are discussed in
\citet{hernan2018causal}, and details for estimating these regressions in the
framework of targeted minimum loss-based estimation are discussed in
\citet{vdl2011targeted}.

@@

\begin{itemize}
\item
  A nonparametric variable importance measure (NPVI) \citep{chambaz2012estimation}:
  The effect of continuous-valued exposure or treatment (the observed
  methylation at a target CpG site) on an outcome of interest is estimated,
  controlling for the observed methylation at all other CpG sites in the same
  neighborhood as the target (treatment) site, based on a parameter that
  compares values of the treatment against a reference value taken to be the
  null. In particular, the implementation provided is designed to assess the
  effect of differential methylation at the target CpG site on a (typically)
  phenotype-level outcome of interest (e.g., survival), in effect providing an
  nonparametric evaluation of the impact of methylation at the target site on
  said outcome.
\end{itemize}

\hypertarget{class-methytmle}{%
\paragraph{\texorpdfstring{Class \texttt{methytmle}}{Class methytmle}}\label{class-methytmle}}

@@

We refer the reader to the package vignette, ``\texttt{methyvim}: Targeted Data-Adaptive
Estimation and Inference for Differential Methylation Analysis,'' included in any
distribution of the software package, for further details.

\hypertarget{operation}{%
\subsection{Operation}\label{operation}}

@@

\hypertarget{use-cases}{%
\section{Use Cases}\label{use-cases}}

@@

\hypertarget{preliminaries-setting-up-the-data}{%
\paragraph{Preliminaries: Setting up the Data}\label{preliminaries-setting-up-the-data}}

@@

\begin{verbatim}
## methyvim v1.1.0: Targeted Variable Importance for Differential Methylation Analysis
\end{verbatim}

@@

\hypertarget{differential-methylation-analysis}{%
\paragraph{Differential Methylation Analysis}\label{differential-methylation-analysis}}

@@

\begin{verbatim}
##                 lwr_ci      est_ate       upr_ci      var_ate         pval
## cg14008030 -0.11956597 -0.031415962  0.056734049 2.022705e-03 4.848468e-01
## cg20253340 -0.08850637 -0.058866142 -0.029225917 2.286919e-04 9.917426e-05
## cg21870274 -0.09499057 -0.029118982  0.036752609 1.129495e-03 3.862537e-01
## cg17308840 -0.04626018 -0.007152452  0.031955277 3.981191e-04 7.199943e-01
## cg00645010 -0.02697515 -0.013809263 -0.000643373 4.512199e-05 3.980386e-02
## cg27534567  0.06745648  0.115711854  0.163967223 6.061486e-04 2.602936e-06
##            n_neighbors n_neighbors_control max_cor_neighbors
## cg14008030           0                   0                NA
## cg20253340           0                   0                NA
## cg21870274           2                   1         0.9443580
## cg17308840           2                   1         0.9443580
## cg00645010           2                   2         0.5236810
## cg27534567           1                   0         0.9362968
\end{verbatim}

Finally, we may compute FDR-corrected p-values, by applying a modified procedure
for controlling the False Discovery Rate for multi-stage analyses (FDR-MSA)
\citep{tuglus2009modified}. We do this by simply applying the \texttt{fdr\_msa} function.

@@

\hypertarget{visualization-of-results}{%
\paragraph{Visualization of Results}\label{visualization-of-results}}

@@

\begin{center}\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-pvals-raw-1} \end{center}

@@

\begin{center}\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-pvals-fdr-1} \end{center}

@@

\begin{center}\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-volcano-1} \end{center}

@@

\begin{center}\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-heatmap-1} \end{center}

Invoking \texttt{methyheat} in this manner produces a plot of the top sites (\(25\), by
default) based on the raw p-value, using the raw methylation measures in the
plot. This uses the exceptional \texttt{superheat} R package \citep{barter2017superheat}.

\hypertarget{summary}{%
\section{Summary}\label{summary}}

@@

\hypertarget{software-availability}{%
\section{Software availability}\label{software-availability}}

@@

\hypertarget{author-contributions}{%
\section{Author contributions}\label{author-contributions}}

NH designed and implemented the software package, applied the tool to the use
cases presented, and co-drafted the present manuscript. RP helped in designing
the software and co-drafted the present article. AH and ML served as advisors
for the development of this software and general statistical algorithm.

\hypertarget{competing-interests}{%
\section{Competing interests}\label{competing-interests}}

@@

\hypertarget{grant-information}{%
\section{Grant information}\label{grant-information}}

NH was supported in part by the National Library of Medicine of the National
Institutes of Health under Award Number T32-LM012417, by P42-ES004705, and
by R01-ES021369. RP was supported by P42-ES004705. The content of this work is
solely the responsibility of the authors and does not necessarily represent the
official views of the various funding sources and agencies.

\renewcommand\refname{References}
{\small\bibliography{paper\_BiocF1000.bib}}
